The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-11-01 Paper.qmd
Paper: Added abstract and conclusion. Edits to discussion and methods.
Regularize: Added coded for full correlation matrix.

@@

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@davison2021; @davison2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@xx, @yy].

@@

Temper tantrums may also be adaptive strategies by which children can elicit more support from parents. They are common in children from 18 months to 4 years old [@potegal2003], and similar behaviors have also been reported in other young animals, often surrounding feeding or weaning conflict [@altmann1980; @goodall1986; @schaller1964]. Temper tantrums vary in their severity, ranging from crying, shrieking, and screaming to violent bodily movements, aggression, and self-harm [@potegal2003]. They have been reported in many societies with different subsistence systems, political organization, and norms surrounding anger and parenting [@albino1956; @banisalameh2021; @fouts2005; @henderson1966; @hilger1957; @konner1974; @ward1970], suggesting they may be a human universal. As with the other proposed signals of need, they are capable of eliciting a wide range of responses, including pro-child responses, neutral responses (e.g., parents ignoring the child) [@hilger1957; @konner1974], and negative responses like abuse [@schmitt1987].

@@

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothesized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children, especially given the lack of prior studies on signaling by older children in low income settings.For this reason, all analyses of these unique data were exploratory and meant to lay the foundation for future studies of child signaling.

@@

XXXXXXXXCHIld RecruitmentXXXXXXXXXXX

@@

All participants for the survey portion of the study met the inclusion criteria. Most adult participants were the biological mothers of one or more children in the study. All children in the study, whose ages ranged from 5-20, were cared for by one of the adult participants.

@@

All adult participants provided informed consent. The surveys, which included many topics outside of child signaling, were designed to last about an hour, plus 20 minutes for each additional child. Adults were paid \$XXXXX for the survey portion of the study.

xxxxxxxxxxxxxChild consent infoxxxxxxxxxxxxxx.

@@

The survey also included many demographic questions about the caretakers and their households. Those used in this study included the (1) ages of those in the household, (2) sexes of those in the household, (3) kinship ties within households, (4) partnership status of the primary caregiver, (5) number of adults who provide childcare within the household, (6) highest level of education of the primary caregiver, (7) immigration status (in regards to Utila), (8) monthly household income, (9) neighborhood of residence, (10) current job of the primary caretaker, and (11) ownership of various possesions (e.g., homes, washing machines, or cellphones; see SIXXXXXX for full list). These data allowed us to calculate the number of children (total, younger, and older) and adults within the household as well as the genetic relatedness between siblings. We also imputed data for caregiver age by setting the missing value to the mean. For the data used in analyses, this amounted to imputing data for one mother which affected two children.

\[mention Campanado here and in ethnographic portion?\]

\[I think this gets deleted or do we mention no findings worth reporting?\] \## Child anthropometrics

@@

<!-- [NOTE: Add categorical variables to the table; combine child and adult variables into 1 table] -->

@@

kable(custom.summarize(modeldf, longform_dict[1:31])) %>% kable_styling()
```

@@

Correlation matrix of the frequencies of signals/states covered in the study with sex, child age, and the frequency of conflict between children and primary caretakers. For a more complete correlation matrix of study variables see SIxxxxxxx.
:::

@@

The effects of particularly large risk or protective factors for child signaling are shown in @fig-signal-effects.

@@

@@

*Negative association between age and child signaling (all signaling measures)*

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003] (check for others), the frequencies of crying and temper tantrums decreased with age, as did the frequency of child sadness, summed signal frequency, and signal cost. One possible reason for this is that as children mature, they are more capable of fulfilling their own desires, more able to ask others for what they want, and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while crying may be costly among the youngest children through its energetic costs, these costs are likely insignificant among older children, who might instead face costs in the form of reputational harm from crying in some contexts but not others. In times when the older child's valuation of potential increased support is small, such costs may be prohibitive. However, in times when the child places more value on potential support, those same costs could allow for a child to credibly signal need if they are too high to be worth it for deceptive signaling. 
- Increased use of signals not mentioned here (.e.g, depression, suicality, others?)

~~Another is that since older children require less investment to reach adulthood than younger children, current or future, they may experience less severe conflicts of interest with  parents. For this reason, less costly signals may be more likely to work to increase investment, all else being equal\[\[4\]\].~~

*Positive association between conflict with parent and child signaling (all signaling measures)*

@@

*Sex and focal child alloparenting effort interact (positive for females, negative/flat for males)*

\[still need the plot for this\]

@@

*Positive associations between child age and signaling frequency and child neediness*

\[Am I missing anything about the significance of this being a relative measure. Could faking drive up the costs in ways everyone signals similarly\]

@@

*Parental responses to last signal of need*

@@

\[delete?\] *In most models, there is a negative association between education and child signaling?*

We found a negative association between caretakers’ education level (years in school) and many of our child signaling measures despite our models controlling for household income and partner status. One possible reason for this may be that more educated caretakers see their children less during the day due to greater ability to afford childcare or due to them having jobs that are less compatible with child care (test model of social support predicted by education). Unfortunately, we do not have data on how much time caretakers spent with children to test this, and it seems equally likely that less educated parents may see their kids less often if they need to work more hours to support their families. Another possibility is that lower levels of education are associated with greater levels of adversity for both parents and children, something which appears to be the case for other measures of SES [@pepper2017] and would increase the benefits of signaling need for children and the difficulty of parents to respond. It is also possible that education influences parenting norms in ways that cause differences in child signaling behavior.

*Demographic effects*

@@

One reason for this might be that neighborhood quality affects the amount of adversity children face, thereby increasing their neediness.

@@

Outside of child anthropometric measurements, all data come from caregiver reports, and therefore, reflect the biases of caretakers, whatever they may be. Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we also faced a tradeoff between survey time and question depth. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.

@@

::: {#stbl-possession-list}
Motorized vehicle (scooter, 4-wheeler, etc.) Bicylce Television (TV) with cable Smartphone Washing machine Refrigerator A home that you own Credit card

@@

::: {#sfig-effect-relativeneed}
```{r, fig.width= 10}
p_need_age_cost
```

The proportion of children perceived by caretakers as being more needy, as needy, or less needy than a caretaker's other children. A: Effect of child age on relative neediness B: Effect of signal cost on relative neediness. Outcome variable question wording: How much time, effort, and resources does this child need compared to your other children?
:::

@@

::: {#sfig-plot-need-coefs}
![](Figures/plot_need_coefs.svg)
Effects of predictor variables on relative need between household children. Ordinal logistic regression model with elastic net penalties.
:::

::: {#sfig-plot-invest-coefs}
![](Figures/plot_invest_coefs.svg)
Effects of predictor variables on relative investment in household children. Ordinal logistic regression model with elastic net penalties.
:::

@@

![](Figures/plot_caregiverresponse_coefs.svg)
Effects of predictor variables on caregiver responses to the last time a child was sad or cried. Ordinal logistic regression model with elastic net penalties.
:::

@@

>>>>

# Abstract

To gain support, children use signals to communicate their desires to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations. However, the full range of potential child signals of need, which extend beyond cries, has never been investigated in children of all ages. To address this, we used exploratory analyses on caregiver reports of child signaling data from xxx families (xx children) on Utila, Honduras. We found that child signaling decreased with age and neighborhood quality and was highly correlated with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were perceived as needier within the household were more likely to receive more investment. They were also more likely to receive a pro-child response when the cause of their last instance of signaling involved discomfort, pain, injury, or, illness, but less likely to receive a beneficial response when it involved prior punishment. Our results show there is much potential to the further study of sadness, crying, and temper tantrums among older children using signaling theory.

@@

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@davison2021; @davison2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@helfrecht2016; @lawson2016; hagen2009].

@@

Temper tantrums may also be adaptive strategies by which children can elicit more support from parents. They are common in children from 18 months to 4 years old [@potegal2003] yet are understudied among older children [@koch2003] outside of studies examining disruptive mood regulation disorder [for one exception see @book]. Temper tantrums vary in their severity, ranging from crying, shrieking, and screaming to violent bodily movements, aggression, and self-harm [@potegal2003], with similar behaviors having also been reported in other young animals, often surrounding feeding or weaning conflict [@altmann1980; @goodall1986; @schaller1964]. Tantrums have been reported in many societies with different subsistence systems, political organization, and norms surrounding anger and parenting [@albino1956; @banisalameh2021; @fouts2005; @henderson1966; @hilger1957; @konner1974; @ward1970], suggesting they may be a human universal. As with the other proposed signals of need, they are capable of eliciting a wide range of responses, including pro-child responses, neutral responses (e.g., parents ignoring the child) [@hilger1957; @konner1974], and negative responses like abuse [@schmitt1987].

@@

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothesized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children, especially given the lack of prior studies on signaling by older children in low income settings. For this reason, all analyses of these unique data were exploratory and meant to lay the foundation for future studies of child signaling.

@@

[no longer needed] XXXXXXXXCHIld RecruitmentXXXXXXXXXXX

@@

All participants for the survey portion of the study met the inclusion criteria. Most ~~adult~~ participants were the biological mothers of one or more children in the study. ~~All children in the study, whose ages ranged from 5-20, were cared for by one of the adult participants.~~

@@

All ~~adult~~ participants provided informed consent. The surveys, which included many topics outside of child signaling, were designed to last about an hour, plus 20 minutes for each additional child. ~~Adults~~ Participants were paid \$XXXXX for the survey portion of the study.

~~xxxxxxxxxxxxxChild consent infoxxxxxxxxxxxxxx.~~

@@

The survey also included many demographic questions about the caretakers and their households. Those used in this study included the (1) ages of those in the household, (2) sexes of those in the household, (3) kinship ties within households, (4) partnership status of the primary caregiver, (5) number of adults who provide childcare within the household, (6) highest level of education of the primary caregiver, (7) immigration status (in regards to Utila), (8) monthly household income, (9) neighborhood of residence, (10) current job of the primary caretaker, and (11) ownership of various possessions (e.g., homes, washing machines, or cellphones; see @stbl-possession-list for full list). These data allowed us to calculate the number of children (total, younger, and older) and adults within the household as well as the genetic relatedness between siblings. We also created a variable for residence in Camponado, a densely populated neighborhood on the island with relatively poor sanitation in which most individuals from the mainland live that is seen as the most dangerous and poorest part of Utila [@garcia2017].

In one instance where data for caregiver age was missing, we set the missing value to the mean. This affected data for two children.

@@

@@

kable(custom.summarize(modeldf, longform_dict[1:23])) %>% kable_styling()
```

@@

Correlation matrix of the frequencies of signals/states covered in the study with sex, child age, and the frequency of conflict between children and primary caretakers. For a more complete correlation matrix of study variables see @sfig-plot-full-corrmat.
:::

@@

The effects of particularly large risk or protective factors for child signaling are shown in @fig-signal-effects and @fig-alloparentingXsex.

@@

::: {#fig-alloparentingXsex}
![](Figures/plot_alloparentingXsex.svg)

Effects of alloparenting frequency (per month) and signal cost and frequency of temper tantrums by sex.
:::

@@

Caregivers reported a wide range of child signaling frequencies. About half of children were reported to cry or throw tantrums more than once a month and slightly less than 40% were reported to engage in them more than once a week. Although similarly common overall, tantrums more common than crying among older children and might be a under-appreciated way by which children might signal for increased support based on these findings and a longitudinal study of temper tantrums and otherchild behaviors which found 50% of boys 30% of girls engaged in temper tantrums at age 9 among a representative, non-clincial sample of 252 children in Berkeley, California [@@macfarlane1962].

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003; @macfarlane1962], the frequencies of crying and temper tantrums decreased with age, as did the frequency of child sadness, summed signal frequency, and signal cost. One possible reason for this is that as children mature, they are more capable of fulfilling their own desires, more able to ask others for what they want, and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while crying may be costly among the youngest children through its energetic costs, these costs are likely insignificant among older children, who might instead face costs in the form of reputational harm from crying in some contexts but not others. In times when an older child's valuation of potential increased support is small, such costs may be prohibitive. However, in times when the child places more value on potential support, those same costs could allow for a child to credibly signal need if they are too high to be worth it for deceptive signaling.

@@

@@

@@

@@

\[delete this paragraph?\] 
We found a negative association between caretakers’ education level (years in school) and many of our child signaling measures despite our models controlling for household income and partner status. One possible reason for this may be that more educated caretakers see their children less during the day due to greater ability to afford childcare or due to them having jobs that are less compatible with child care (test model of social support predicted by education). Unfortunately, we do not have data on how much time caretakers spent with children to test this, and it seems equally likely that less educated parents may see their kids less often if they need to work more hours to support their families. Another possibility is that lower levels of education are associated with greater levels of adversity for both parents and children, something which appears to be the case for other measures of SES [@pepper2017] and would increase the benefits of signaling need for children and the difficulty of parents to respond. It is also possible that education influences parenting norms in ways that cause differences in child signaling behavior.

@@

@@

~~Outside of child anthropometric measurements,~~ All data come from caregiver reports, and therefore, reflect the biases of caretakers, whatever they may be. Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we also faced a tradeoff between survey time and question depth. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.

# Conclusion

Children face the challenge of eliciting support from others in times of adversity and opportunity. The signaling strategies of Utilian children were highly correlated and engaged in by children of all ages. Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals, 1) child signaling effort was positively correlated with the frequency of conflict with caretakers and 2) greater perceptions of child need predicted greater investment in a given child. Our results suggest there is much potential for signaling theory to help explain patterns of child crying and temper tantrums.

@@

::: {#stbl-possession-list}
Motorized vehicle (scooter, 4-wheeler, etc.) Bicycle Television (TV) with cable Smartphone Washing machine Refrigerator A home that you own Credit card

@@

::: {#sfig-plot-full-corrmat}
![](Figures/plot_fullcorrmat.svg)

Complete correlation matrix of study variables.
:::

@@

::: {#sfig-plot-need-coefs}
![](Figures/plot_need_coefs.svg)

Effects of predictor variables on relative need between household children. Ordinal logistic regression model with elastic net penalties.
:::

::: {#sfig-plot-invest-coefs}
![](Figures/plot_invest_coefs.svg)

Effects of predictor variables on relative investment in household children. Ordinal logistic regression model with elastic net penalties.
:::

@@

![](Figures/plot_caregiverresponse_coefs.svg)

Effects of predictor variables on caregiver responses to the last time a child was sad or cried. Ordinal logistic regression model with elastic net penalties.
:::

@@

### Child anthropometrics

### Old discussion parts
